---
title: "From an inference result to a grain plan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From an inference result to a grain plan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~1.5 s under rmarkdown::render(); macOS arm64, R 4.5.2, one
     core; agridat, ggplot2 and knitr all installed on the measured build. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  dpi = 150
)
set.seed(1L)
```

```{r library, message = FALSE}
library(grainPlan)
have_agridat <- requireNamespace("agridat", quietly = TRUE)
have_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

An agronomist closing out a season has three inference results waiting on
the desk -- a nitrogen-response trial, a variety trial, a protein target --
and one blunt question for each: *what should I actually do about it, and
how do I tell when the tool itself does not trust the evidence enough to
answer?* A wrong number costs a season's margin; a confident wrong number is
worse than a plain shrug, because nobody double-checks a number that looks
sure of itself.

grainPlan is the practitioner last-mile of the ORCHESTRA agricultural
analytics stack. An upstream member produces an inference result -- a yield
posterior, a multi-environment variety trial, a genomic prediction -- and
grainPlan turns it into an actionable grain-production plan: how much
nitrogen to apply, which variety to sow, whether to chase a quality grade.
It is a thin grain-semantics layer over the `decideR` decision engine: the
expected-utility maximisation, the abstention logic, and the grounding
firewall are all `decideR`'s, reused whole; grainPlan adds the grain
economics and the grain meaning. This walkthrough runs end-to-end on real
public data from the *agridat* package and answers the agronomist's
question three times over -- a rate, a variety, a grade target -- and then
a fourth time for the case a report cannot afford to get wrong: what the
plan looks like when one of its legs is not there to trust.

## What

Four decision verbs and two orchestrating types carry the whole of this
vignette. `plan_nitrogen_rate()` and `plan_nitrogen_rate_from_manifest()`
price a nitrogen rate by expected profit; `plan_variety()` and
`plan_variety_from_manifest()` rank a candidate variety or hybrid under a
downside-aware merit loss; `plan_grade_target()` chooses how hard to push a
quality-driving input under a payoff that steps at a delivery-grade
threshold. Every one of these wraps `decideR::decide()` or a sibling engine
call and returns a `grain_decision`: the recommended action, its unit, a
grower-facing rationale, and a grounding token copied straight off the
wrapped `decideR::decision`, never recomputed. `plan_season()` gathers
several `grain_decision`s into one `grain_plan`, combining their grounding
worst-case through `decideR::combine_grounding()` so a plan resting on any
unverified decision is itself unverified; `grain_plan_table()` flattens that
plan into the one-row-per-decision table a report reads, and
`grain_is_grounded()` is the predicate a downstream automation gates on
before acting. The grain value library -- `wheat_protein_bands()`,
`barley_malting_premium()`, `n_rate_economics()` -- consolidates the
economics these decisions price against, so a schedule is named and
documented rather than an anonymous expression re-derived at every call
site.

## Do

### A nitrogen rate from a real yield response

```{r skip-nitrogen, echo = FALSE, eval = !have_agridat, results = "asis"}
cat("agridat is not installed on this build, so the nitrogen-rate example",
    "below does not run.\n")
```

The *agridat* dataset `lasrosas.corn` records corn yield-monitor data for a
field near Las Rosas, Rio Cuarto, Cordoba, Argentina, under variable
nitrogen fertiliser across the 1999 and 2001 harvests (Anselin, Bongiovanni
& Lowenberg-DeBoer, 2004). We summarise it into a yield-response curve,
build a posterior of yield at each candidate rate, and ask grainPlan for
the profit-optimal rate. The economics live in the grain value library:
`n_rate_economics()` consolidates the profit utility -- revenue at the
grain price less the nitrogen cost -- so a profit-priced rate decision reads
as its economics rather than as an anonymous expression.

```{r nitrogen, eval = have_agridat}
data("lasrosas.corn", package = "agridat")
corn <- lasrosas.corn

# A candidate nitrogen grid spanning the trial, and a simple yield-response
# summary: a quadratic mean-yield fit against nitrogen, with the trial's
# residual spread used to build a posterior of yield at each rate. This
# stands in for the yield posterior an ORCHESTRA inference member (a PESTO
# inversion, a TACI yield-response model) would emit as an
# orchestra_manifest.
#
# The draws MUST be coupled across rate columns: one draw index has to mean
# "the same realised season" at every candidate rate, because the decision
# is a paired comparison between candidate rates on the posterior, not a
# comparison of two independent samples. We therefore draw one residual
# vector `eps` and add it to every rate's fitted mean, rather than drawing
# each column independently (independent columns roughly double the
# between-rate contrast variance and can mask a real, decisive difference).
rates <- seq(0, 120, by = 20)
fit <- stats::lm(yield ~ nitro + I(nitro^2), data = corn)
sigma <- summary(fit)$sigma

n_draws <- 1500L
eps <- stats::rnorm(n_draws, 0, sigma)
yield_draws <- vapply(
  rates,
  function(r) {
    mu <- stats::predict(fit, newdata = data.frame(nitro = r))
    mu + eps
  },
  numeric(n_draws))

# Grain and nitrogen prices are the caller's to supply. `lasrosas.corn`
# reports yield in quintals/ha (agridat's own documentation), so the grain
# price below is expressed per quintal; it is illustrative, not the price
# structure the trial's own economic analyses used (Anselin et al. (2004)
# report $6.85/quintal and $0.4348/kg N for their own analysis).
n_plan <- plan_nitrogen_rate(
  yield_draws, rates,
  price_grain = 22,             # $ per quintal (illustrative)
  price_n = 1.2,                # $ per unit nitrogen (illustrative)
  crop = "corn",
  grounding = decideR::grounding_grounded())

n_plan
```

The recommendation is a `grain_decision`: the chosen rate, the crop, a
grower-facing rationale, and -- carried beside the numbers, never inferred
from them -- the grounding token.

```{r fig-nitrogen, eval = have_agridat && have_ggplot2, echo = have_agridat && have_ggplot2, fig.cap = "Expected profit against candidate nitrogen rate for the lasrosas.corn fit: profit rises from the status quo, peaks at the chosen rate, and falls away beyond it."}
plot_df <- n_plan@decision@candidates
plot_df$role <- "candidate"
plot_df$role[plot_df$action == 0] <- "status quo"
plot_df$role[plot_df$action == n_plan@action] <- "chosen rate"
plot_df$role <- factor(plot_df$role,
                       levels = c("candidate", "status quo", "chosen rate"))

ggplot2::ggplot(plot_df, ggplot2::aes(x = action, y = expected_utility)) +
  ggplot2::geom_line(colour = "grey50") +
  ggplot2::geom_point(ggplot2::aes(colour = role, shape = role), size = 3) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c(candidate = "#000000", `status quo` = "#0072B2",
               `chosen rate` = "#D55E00")) +
  ggplot2::scale_shape_manual(
    name = NULL,
    values = c(candidate = 16, `status quo` = 15, `chosen rate` = 17)) +
  ggplot2::labs(
    x = "Nitrogen rate (kg N/ha)", y = "Expected profit ($/unit area)",
    title = "Profit-vs-rate curve behind the recommendation") +
  ggplot2::theme_minimal()
```

```{r fig-nitrogen-skip, echo = FALSE, eval = !(have_agridat && have_ggplot2), results = "asis"}
cat("agridat and/or ggplot2 is not installed on this build, so the",
    "profit-vs-rate figure is skipped.\n")
```

The curve is why the decision is decisive rather than a coin toss between
neighbouring rates: expected profit rises steadily from the status quo,
peaks sharply at the chosen rate, and falls away either side of it, so the
gap between the chosen rate and its nearest neighbours is well above the
posterior's rate-to-rate noise.

### The firewall: unverified evidence abstains

Now the decisive behaviour. If the yield evidence is not grounded -- the
upstream producer did not verify it against an external authority -- the
same decision must not be made. grainPlan inherits `decideR`'s Independent
Oracle Principle firewall: an unverified input forces abstention to the
status-quo rate, never a confident wrong recommendation.

```{r firewall, eval = have_agridat}
n_plan_unverified <- plan_nitrogen_rate(
  yield_draws, rates,
  price_grain = 22, price_n = 1.2, crop = "corn",
  grounding = decideR::grounding_unverified())

n_plan_unverified
```

The only thing that changed between this call and the one above is the
grounding token, and it flipped decide into abstain. This is the firewall a
real grain tool needs: a nitrogen rate carries a cost in the paddock, so an
unverified yield model is held at the status quo, not acted on.

### A variety choice from a multi-environment trial

```{r skip-variety, echo = FALSE, eval = !have_agridat, results = "asis"}
cat("agridat is not installed on this build, so the variety-choice example",
    "below does not run.\n")
```

The *agridat* dataset `besag.met` is a multi-environment trial of 64 corn
hybrids across six counties in North Carolina, with three replicates each
(Besag & Higdon, 1999). We summarise each hybrid's mean performance across
counties into a posterior of merit, and ask grainPlan to rank them under a
downside-aware loss -- one that discounts a high-yielding but volatile
hybrid against a steady one, the stability a grower wants from a sown
variety.

```{r variety, eval = have_agridat}
data("besag.met", package = "agridat")
met <- besag.met

# Hybrid means across counties, with the county-to-county spread giving each
# hybrid a merit posterior. (A factor-analytic MET model --
# flexyBayes -- would emit this directly as an orchestra_manifest;
# here we build it directly from the public trial.) The posterior we need
# is of the HYBRID MEAN, not of a single plot, so the per-hybrid plot-level
# spread must be divided down to a standard error of the mean
# (sd / sqrt(n)); using the raw plot-level sd would overstate the mean's
# uncertainty by roughly sqrt(n) and could mask a real, decisive ranking.
# The trial ships with 36 filler rows (six per county) carrying a missing
# yield -- agridat's own note explains they complete the field rectangle for
# a spatial model and were never harvested -- so we summarise with na.rm and
# fall back to the trial-wide spread (also converted to a standard error)
# for any hybrid with too few records to estimate its own.
gen_means <- tapply(met$yield, met$gen, mean, na.rm = TRUE)
gen_sd <- tapply(met$yield, met$gen, stats::sd, na.rm = TRUE)
gen_n <- tapply(met$yield, met$gen, function(x) sum(!is.na(x)))
pooled_sd <- stats::sd(met$yield, na.rm = TRUE)
gen_sd[is.na(gen_sd)] <- pooled_sd
gen_n[gen_n < 2L] <- 1L
gen_se <- gen_sd / sqrt(gen_n)
varieties <- names(gen_means)

merit_draws <- vapply(
  seq_along(gen_means),
  function(j) stats::rnorm(1000L, gen_means[j], gen_se[j]),
  numeric(1000L))
colnames(merit_draws) <- varieties

v_plan <- plan_variety(
  merit_draws, varieties, risk_aversion = 0.5, crop = "corn",
  grounding = decideR::grounding_grounded())

v_plan
```

The recommendation is the leading hybrid on the downside-aware ranking; the
full ranked short-list is in the wrapped decision's candidate ledger, with
the hybrid labels attached (the ledger's `action` column carries the
integer index grainPlan decides over, so we join the hybrid names back on
for a readable table):

```{r variety-shortlist, eval = have_agridat}
ledger <- v_plan@decision@candidates
ledger$hybrid <- varieties[ledger$action]
ledger <- ledger[order(-ledger$expected_utility),
                 c("hybrid", "action", "expected_utility", "feasible")]
```

```{r variety-shortlist-kable, echo = FALSE, eval = have_agridat}
knitr::kable(
  head(ledger, 5L), digits = 2, row.names = FALSE,
  col.names = c("Hybrid", "Index", "Expected merit", "Feasible"),
  caption = "The five leading corn hybrids on the downside-aware merit ranking."
)
```

### A quality-grade decision

Many grain payoffs are not smooth in the outcome they reward: a delivery is
paid by quality grade, so the value steps at a protein threshold rather
than rising continuously. grainPlan's value library consolidates these
grade schedules. `wheat_protein_bands()` prices the Australian
milling-wheat grades by protein, following the band thresholds published by
Grain Trade Australia, and `plan_grade_target()` decides how hard to push a
protein-driving input under that step payoff -- a decision governed by the
posterior probability of clearing each band, exactly the threshold
economics `decideR`'s grade-band loss expresses.

```{r grade-schedule}
protein_value <- wheat_protein_bands()
```

```{r grade-schedule-kable, echo = FALSE}
knitr::kable(
  data.frame(band = attr(protein_value, "labels"),
             payoff = attr(protein_value, "values")),
  col.names = c("Band", "Payoff ($/tonne)"),
  caption = paste0(
    "Wheat protein-band payoffs (`wheat_protein_bands()` defaults; dollar ",
    "values are illustrative, the band shape follows Grain Trade ",
    "Australia's published thresholds).")
)
```

```{r fig-protein-bands, eval = have_ggplot2, echo = have_ggplot2, fig.cap = "The wheat protein-band value schedule (`wheat_protein_bands()`): a price per tonne that steps up at each published grade threshold. Dashed lines mark the milling/hard/prime-hard breaks; the step shape is what the table above cannot show."}
grid_protein <- seq(9, 14, by = 0.02)
schedule_df <- data.frame(
  protein = grid_protein,
  payoff = protein_value(grid_protein))
breaks_df <- data.frame(protein = attr(protein_value, "breaks"))

ggplot2::ggplot(schedule_df, ggplot2::aes(x = protein, y = payoff)) +
  ggplot2::geom_step(colour = "#0072B2", linewidth = 0.9) +
  ggplot2::geom_vline(data = breaks_df, ggplot2::aes(xintercept = protein),
                       linetype = "dashed", colour = "grey40") +
  ggplot2::labs(
    x = "Grain protein (%)", y = "Payoff ($/tonne)",
    title = "Wheat protein-band value schedule") +
  ggplot2::theme_minimal()
```

```{r fig-protein-bands-skip, echo = FALSE, eval = !have_ggplot2, results = "asis"}
cat("ggplot2 is not installed on this build, so the protein-band schedule",
    "figure is skipped.\n")
```

A baseline protein posterior just below the milling band, and a nitrogen
top-up that lifts protein at a per-unit cost, gives grainPlan something to
weigh against that schedule: the decision prices the chance of clearing the
11.5% band against the cost of pushing for it.

```{r grade}
protein <- stats::rnorm(2000L, mean = 11.0, sd = 0.5)
g_plan <- plan_grade_target(
  protein, actions = seq(0, 60, by = 10), value = protein_value,
  quality_shift = function(rate, p) p + 0.02 * rate, cost = 0.3,
  crop = "wheat", grounding = decideR::grounding_grounded())

g_plan
```

At this cost the top-up pays for itself: the recommended input level is the
top of the candidate range, because the expected graded payoff rises
monotonically with the higher probability of clearing the 11.5% band -- the
candidate ledger below shows the full trade-off.

```{r grade-candidates, echo = FALSE}
knitr::kable(
  g_plan@decision@candidates, digits = 2, row.names = FALSE,
  col.names = c("Input level", "Expected payoff ($/t)", "Feasible"),
  caption = "Full candidate ledger behind the grade-target recommendation."
)
```

### From an upstream manifest

The examples so far assembled the yield and merit posteriors by hand. In
the orchestra, an upstream member emits an `orchestra_manifest` instead --
the C2 v2 generic manifest contract in the orchestra's cross-package
governance -- and grainPlan reads it directly through `decideR`'s
duck-typed pipeline tail, taking no dependency on the manifest constructor
package. Here is a minimal stand-in built from a tiny S7 class, standing
for the manifest a real producer would emit.

```{r manifest-mini}
mini_manifest <- S7::new_class(
  "mini_manifest",
  properties = list(
    outputs         = S7::class_any,
    metadata        = S7::class_list,
    run_id          = S7::class_character,
    emitter_package = S7::class_character))
```

A yield-per-rate draws matrix, the shape a PESTO inversion emits, carried
in `outputs` with the producer's grounding token in `metadata$grounding`:

```{r manifest-rate}
rates_m <- seq(0, 150, by = 25)
ymax    <- stats::rnorm(1200L, 5.5, 0.3)
yld_m   <- vapply(rates_m,
                  function(r) 3 + (ymax - 3) * (1 - exp(-r / 80)),
                  numeric(1200L))
ym <- mini_manifest(outputs = yld_m,
                    metadata = list(grounding = decideR::grounding_grounded()))

plan_nitrogen_rate_from_manifest(ym, rates_m, price_grain = 350, price_n = 1.3,
                                 crop = "wheat")
```

The genomic tail reads breeding values the same way. A `breeding_values`
manifest carries point GEBVs in `outputs$gebv` alongside their genomic
uncertainty, so the reconstruction uses the genotypes' own reliability
rather than an invented spread.

```{r manifest-gebv}
gm <- mini_manifest(
  outputs  = list(gebv = c(Scepter = 0.2, Vixen = 1.1, Calibre = -0.3),
                  pev  = c(0.04, 0.04, 0.04)),
  metadata = list(grounding = decideR::grounding_grounded()))

g_dec <- plan_variety_from_manifest(gm, risk_aversion = 0.5)
g_dec
g_dec@context$gebv_sd_source
```

If a point-GEBV manifest carries no uncertainty and no `gebv_sd` is
supplied, grainPlan stops rather than fabricate a scale for the
reconstruction: a genotype switch must not turn on a magic constant.

```{r manifest-no-sd, error = TRUE}
gm_no_sd <- mini_manifest(
  outputs  = list(gebv = c(Scepter = 0.2, Vixen = 1.1, Calibre = -0.3)),
  metadata = list(grounding = decideR::grounding_grounded()))

plan_variety_from_manifest(gm_no_sd, risk_aversion = 0.5)
```

The stop is not a bug to route around: it is the same firewall discipline
as the grounding token, applied to a different way an upstream producer's
claim can be too thin to act on.

A manifest can also carry a refusal of its own. When the upstream producer
declares, in its own `summary` slot, that it did not produce a usable
result -- a PESTO ensemble that never converged, say -- grainPlan reads that
typed abstention before it ever looks at the draws, and holds the decision
at the status quo regardless of what the (declined) manifest happens to
carry:

```{r manifest-refusal}
refused_manifest <- mini_manifest(
  outputs = list(),
  metadata = list(
    grounding = decideR::grounding_grounded(),
    summary = list(headline = "no stable fit found", abstained = TRUE)),
  run_id = "demo-refused-01", emitter_package = "PESTO")

gd_refused <- plan_nitrogen_rate_from_manifest(
  refused_manifest, rates_m, price_grain = 350, price_n = 1.3, crop = "corn")

gd_refused
```

Note that the producer's own grounding token claimed `grounded`; grainPlan
never reaches it, because the typed abstention is checked first. The
returned object also carries a class stamp a fleet-level gate can read
without knowing it came from grainPlan specifically:

```{r manifest-refusal-class}
class(gd_refused)
```

`grainPlan_abstention` and `orchestra_refusal` are prepended ahead of
`grainPlan::grain_decision`, the naming convention the orchestra's
cross-member decline gate reads producer-agnostically -- a downstream
automation recognises this as a refused decision without special-casing
grainPlan's own class.

### Composing a season plan

Finally, the orchestrator. `plan_season()` gathers several grain decisions
into one `grain_plan` whose grounding is the worst case across its
decisions -- a plan that rests on any unverified decision is itself
unverified, the firewall applied one level up. A plan with a declined
component is flagged so a report never reads a confident headline over it.

```{r skip-season, echo = FALSE, eval = !have_agridat, results = "asis"}
cat("agridat is not installed on this build, so the season-composition",
    "examples below do not run.\n")
```

The nitrogen and variety decisions above come from two separate real
trials -- a precision-agriculture nitrogen trial in Argentina and a
multi-environment hybrid trial in North Carolina -- composed here only to
show the orchestrator mechanics; a real season plan would draw both
decisions from the one field. Because both trials happen to be corn, the
plan below labels its crop explicitly rather than infer it from either
input, so the composition is always unambiguous.

```{r season, eval = have_agridat}
season <- plan_season(
  list(v_plan, n_plan),
  crop = "corn", season = "2026 demonstration")

season
season_tbl <- grain_plan_table(season)
```

```{r season-kable, echo = FALSE, eval = have_agridat}
knitr::kable(
  season_tbl, digits = 1, row.names = FALSE,
  col.names = c("Decision", "Action", "Unit", "Abstained", "Grounding"),
  caption = paste0(
    "The season plan flattened for a report (plan grounding: ",
    attr(season_tbl, "plan_grounding"), ").")
)
```

Every recommendation in the plan carries its grounding, and the plan as a
whole carries the worst case -- so a grower, or a downstream automation, can
gate on `grain_is_grounded()` before acting.

```{r gate, eval = have_agridat}
grain_is_grounded(season)
```

Both decisions above were grounded, so the plan reads clean end to end.
That is the easy case. The harder, more consequential case is what
happens when one leg of the season inherits a refusal from further
upstream -- the `gd_refused` decision built two sections back, standing in
for a PESTO ensemble that never converged. Composed alongside the same
variety decision, into a plan for the same crop:

```{r season-refusal, eval = have_agridat}
season_refusal <- plan_season(
  list(v_plan, gd_refused),
  crop = "corn",
  season = "2026 demonstration -- nitrogen evidence withheld upstream")

season_refusal
refusal_tbl <- grain_plan_table(season_refusal)
```

```{r season-refusal-kable, echo = FALSE, eval = have_agridat}
knitr::kable(
  refusal_tbl, digits = 1, row.names = FALSE,
  col.names = c("Decision", "Action", "Unit", "Abstained", "Grounding"),
  caption = paste0(
    "The same report over a season where the nitrogen leg inherits an ",
    "upstream producer refusal (plan grounding: ",
    attr(refusal_tbl, "plan_grounding"), ").")
)
```

```{r gate-refusal, eval = have_agridat}
grain_is_grounded(season_refusal)
```

## Read

```{r nitrogen-answer-yes, echo = FALSE, eval = have_agridat, results = "asis"}
cat(sprintf(
  paste0(
    "For the lasrosas.corn fit, grainPlan recommends **%g kg N/ha**, at an ",
    "expected profit of **$%s per unit area**, against a status-quo ",
    "profit of $%s -- the profit gain the figure shows as the gap between ",
    "the vermillion point and the blue square. The recommendation carries ",
    "the `%s` token, so it is the number a grower acts on.\n\n"),
  n_plan@action,
  format(n_plan@decision@expected_utility, digits = 5L),
  format(n_plan@decision@candidates$expected_utility[
    n_plan@decision@candidates$action == 0], digits = 5L),
  as.character(n_plan@grounding)))
```

```{r nitrogen-answer-no, echo = FALSE, eval = !have_agridat, results = "asis"}
cat("The nitrogen-rate example did not run on this build (agridat is not",
    "installed), so its recommended rate and profit are not reported",
    "here.\n\n")
```

```{r variety-answer, echo = FALSE, eval = have_agridat, results = "asis"}
top_row <- ledger[1L, ]
cat(sprintf(
  paste0(
    "On the same trial, the leading hybrid is **%s**, with an expected ",
    "downside-aware merit of **%.2f**; the short-list table carries every ",
    "hybrid the ranking considered, in order.\n\n"),
  top_row$hybrid, top_row$expected_utility))
```

```{r grade-answer, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    "For the protein target, the top-up pays for itself at every rate up ",
    "to the top of the candidate grid: grainPlan recommends an input ",
    "level of **%g**, at an expected graded payoff of **$%s per tonne**.\n\n"
  ),
  g_plan@action, format(g_plan@decision@expected_utility, digits = 5L)))
```

```{r season-answer, echo = FALSE, eval = have_agridat, results = "asis"}
cat(sprintf(
  paste0(
    "The first season plan carries **%d** decisions, both decided, and a ",
    "plan-level grounding of `%s`: `grain_is_grounded()` returns `%s`, so ",
    "a downstream gate lets the whole plan through. The second season plan ",
    "carries the same variety decision beside the withheld nitrogen ",
    "decision; its `Abstained` column reads `TRUE` for the nitrogen row ",
    "and its `Action` column carries **%g kg N/ha** -- the status-quo ",
    "fallback, not a priced recommendation -- while the variety row is ",
    "untouched. The plan's own grounding drops to `%s` and ",
    "`grain_is_grounded()` returns `%s`. A report built on ",
    "`grain_plan_table()` reads exactly those two columns, `Abstained` and ",
    "`Grounding`, to decide how to render each row: a row with ",
    "`Abstained = TRUE` is greyed out and printed as a declined ",
    "recommendation carrying only its status-quo fallback, never presented ",
    "beside a genuine number as though it were one.\n\n"
  ),
  nrow(season_tbl), as.character(season@grounding),
  grain_is_grounded(season), gd_refused@action,
  as.character(season_refusal@grounding),
  grain_is_grounded(season_refusal)))
```

That answers the agronomist's question on all four counts: a rate, a
hybrid, and a grade target when the evidence is grounded, and a plain,
legible refusal -- not a confident number -- when a leg of the season is
not there to trust.

## Limits

The dollar values throughout -- the nitrogen prices, the protein-band
payoffs, the malting premium the vignette does not exercise -- are
illustrative and not current market prices or GRDC-origin figures; a real
decision needs the caller's own price sheet. The nitrogen-response fit
pools both recorded harvest years (1999 and 2001) without a year effect, so
the response curve is an average across two seasons rather than one
season's economics, and the season-composition examples draw their two
decisions from two real but unrelated trials -- an Argentine
precision-agriculture strip trial and a North Carolina hybrid trial -- to
demonstrate the orchestrator's mechanics, not to stand in for one grower's
actual paddock. `plan_season()` in this release composes by sequencing and
worst-case grounding only: it does not re-optimise jointly across
decisions, so a season plan is the sum of its parts, not a jointly optimal
allocation across them. The grounding firewall itself trusts the token a
caller or an upstream manifest supplies; grainPlan, like `decideR`, does
not independently re-verify a producer's claim of `grounded` against an
external authority, it only propagates and combines the token worst-case --
so a producer that mislabels its own evidence would still ride through, and
the firewall protects only against evidence that declares its own
unverified or abstained status, not evidence that misrepresents it.

## What to read next

*Getting started with decideR* walks the expected-utility engine, the
abstention logic and the grounding firewall from the ground up -- every
verb in this vignette wraps that engine rather than reimplementing it.

## Reproduce

Seed `1L`, set once at the top of this file; every draw in every section
follows deterministically from that one stream.

```{r session-info}
sessionInfo()
```
