---
title: "From an inference result to a grain plan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From an inference result to a grain plan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1L)
```

grainPlan is the practitioner last-mile of the ORCHESTRA agricultural analytics
stack. An upstream member produces an inference result -- a yield posterior, a
multi-environment variety trial, a genomic prediction -- and grainPlan turns it
into an actionable grain-production plan: how much nitrogen to apply, which
variety to sow, whether to chase a quality grade. It is a thin grain-semantics
layer over the `decideR` decision engine: the expected-utility maximisation, the
abstention logic, and the grounding firewall are all `decideR`'s, reused whole;
grainPlan adds the grain economics and the grain meaning.

This walkthrough runs end-to-end on public data from the *agridat* package -- a
real nitrogen-response trial and a real variety trial -- and shows the two
behaviours that matter for a consequential real-world tool: grounded evidence
*decides*, and unverified evidence *abstains* to the status quo.

```{r libs, message = FALSE}
library(grainPlan)
have_agridat <- requireNamespace("agridat", quietly = TRUE)
```

## A nitrogen rate from a real yield response

The *agridat* dataset `lasrosas.corn` records yield and applied nitrogen across
a precision-agriculture corn trial near Las Rosas, Argentina. We summarise it
into a yield-response curve, build a posterior of yield at each candidate rate,
and ask grainPlan for the profit-optimal rate.

The economics live in the grain value library. `n_rate_economics()` is the
consolidation of the profit utility the orchestra runs kept hand-building --
revenue at the grain price less the nitrogen cost -- so a profit-priced rate
decision reads as its economics rather than as an anonymous expression.

```{r nitrogen, eval = have_agridat}
data("lasrosas.corn", package = "agridat")
corn <- lasrosas.corn

# A candidate nitrogen grid spanning the trial, and a simple yield-response
# summary: mean yield in each nitrogen neighbourhood, with the trial's residual
# spread used to build a posterior of yield at each rate. This stands in for the
# yield posterior an ORCHESTRA inference member (a PESTO inversion, a TACI
# yield-response model) would emit as an orchestra_manifest.
#
# The draws MUST be coupled across rate columns: one draw index has to mean
# "the same realised season" at every candidate rate, because the decision
# is a *paired* comparison between candidate rates on the posterior, not a
# comparison of two independent samples. We therefore draw one residual
# vector `eps` and add it to every rate's fitted mean, rather than drawing
# each column independently (independent columns roughly double the
# between-rate contrast variance and can mask a real, decisive difference).
rates <- seq(0, 120, by = 20)
fit <- stats::lm(yield ~ nitro + I(nitro^2), data = corn)
sigma <- summary(fit)$sigma

n_draws <- 1500L
eps <- stats::rnorm(n_draws, 0, sigma)
yield_draws <- vapply(
  rates,
  function(r) {
    mu <- stats::predict(fit, newdata = data.frame(nitro = r))
    mu + eps
  },
  numeric(n_draws))

# Grain and nitrogen prices are the caller's to supply. `lasrosas.corn` reports
# yield in quintals/ha (agridat's own documentation), so the grain price below
# is expressed per quintal; it is illustrative, not a grounded market quote.
n_plan <- plan_nitrogen_rate(
  yield_draws, rates,
  price_grain = 22,             # $ per quintal (illustrative)
  price_n = 1.2,                # $ per unit nitrogen (illustrative)
  crop = "corn",
  grounding = decideR::grounding_grounded())

n_plan
```

The recommendation is a `grain_decision`: the chosen rate, the crop, a
grower-facing rationale, and -- carried beside the numbers, never inferred from
them -- the grounding token.

```{r nitrogen-fig, eval = have_agridat, fig.cap = "Expected profit against candidate nitrogen rate for the lasrosas.corn fit: profit rises from the status quo, peaks at the chosen rate, and falls away beyond it."}
candidates <- n_plan@decision@candidates
graphics::plot(
  candidates$action, candidates$expected_utility,
  type = "b", pch = 16,
  xlab = "Nitrogen rate (kg N/ha)", ylab = "Expected profit ($/unit area)",
  main = "Profit-vs-rate curve behind the recommendation")
graphics::points(n_plan@action, max(candidates$expected_utility),
                  pch = 17, col = "forestgreen", cex = 1.4)
graphics::points(0, candidates$expected_utility[candidates$action == 0],
                  pch = 15, col = "grey40", cex = 1.2)
graphics::legend("bottomright", bty = "n",
  legend = c("chosen rate", "status quo (0 kg N/ha)"),
  pch = c(17, 15), col = c("forestgreen", "grey40"))
```

The curve is why the decision is decisive rather than a coin toss between
neighbouring rates: expected profit rises steadily from the status quo,
peaks sharply at the chosen rate, and falls away either side of it, so the
gap between the chosen rate and its nearest neighbours is well above the
posterior's rate-to-rate noise.

### The firewall: unverified evidence abstains

Now the decisive behaviour. If the yield evidence is *not* grounded -- the
upstream producer did not verify it against an external authority -- the same
decision must not be made. grainPlan inherits `decideR`'s Independent Oracle
Principle firewall: an unverified input forces an honest abstention to the
status-quo rate, never a confident wrong recommendation.

```{r firewall, eval = have_agridat}
n_plan_unverified <- plan_nitrogen_rate(
  yield_draws, rates,
  price_grain = 0.25, price_n = 1.2, crop = "corn",
  grounding = decideR::grounding_unverified())

n_plan_unverified
```

The only thing that changed is the grounding token, and it flipped *decide* into
*abstain*. This is the firewall a real grain tool needs: a nitrogen rate carries
a cost in the paddock, so an unverified yield model is held at the status quo,
not acted on.

## A variety choice from a multi-environment trial

The *agridat* dataset `besag.met` is a multi-environment variety trial: 64
genotypes across six environments. We summarise each genotype's mean performance
across environments into a posterior of merit, and ask grainPlan to rank them
under a downside-aware loss -- one that discounts a high-yielding but volatile
line against a steady one, the stability a grower wants from a sown variety.

```{r variety, eval = have_agridat}
data("besag.met", package = "agridat")
met <- besag.met

# Genotype means across environments, with the trial's environment-to-environment
# spread giving each genotype a merit posterior. (A factor-analytic MET model --
# flexyBayes::fb_met_summary -- would emit this as an orchestra_manifest; here we
# build it directly from the public trial.) The posterior we need is of the
# GENOTYPE MEAN, not of a single plot, so the per-genotype plot-level spread
# must be divided down to a standard error of the mean (sd / sqrt(n)); using
# the raw plot-level sd would overstate the mean's uncertainty by roughly
# sqrt(n) and could mask a real, decisive ranking. A few yield cells are
# missing, so we summarise with na.rm and fall back to the trial-wide spread
# (also converted to a standard error) for any genotype with too few records
# to estimate its own.
gen_means <- tapply(met$yield, met$gen, mean, na.rm = TRUE)
gen_sd <- tapply(met$yield, met$gen, stats::sd, na.rm = TRUE)
gen_n <- tapply(met$yield, met$gen, function(x) sum(!is.na(x)))
pooled_sd <- stats::sd(met$yield, na.rm = TRUE)
gen_sd[is.na(gen_sd)] <- pooled_sd
gen_n[gen_n < 2L] <- 1L
gen_se <- gen_sd / sqrt(gen_n)
varieties <- names(gen_means)

merit_draws <- vapply(
  seq_along(gen_means),
  function(j) stats::rnorm(1000L, gen_means[j], gen_se[j]),
  numeric(1000L))
colnames(merit_draws) <- varieties

v_plan <- plan_variety(
  merit_draws, varieties, risk_aversion = 0.5, crop = "wheat",
  grounding = decideR::grounding_grounded())

v_plan
```

The recommendation is the leading genotype on the downside-aware ranking; the
full ranked short-list is in the wrapped decision's candidate ledger, with the
genotype labels attached (the ledger's `action` column carries the integer
index grainPlan decides over, so we join the variety names back on for a
readable table):

```{r shortlist, eval = have_agridat}
ledger <- v_plan@decision@candidates
ledger$variety <- varieties[ledger$action]
ledger <- ledger[order(-ledger$expected_utility), c("variety", "action", "expected_utility", "feasible")]
head(ledger, 5L)
```

## A quality-grade decision

Many grain payoffs are not smooth in the outcome they reward: a delivery is paid
by quality *grade*, so the value steps at a protein threshold rather than rising
continuously. grainPlan's value library consolidates these grade schedules.
`wheat_protein_bands()` prices the Australian milling-wheat grades by protein,
and `plan_grade_target()` decides how hard to push a protein-driving input under
that step payoff -- a decision governed by the posterior probability of clearing
each band, exactly the threshold economics `decideR`'s grade-band loss expresses.

```{r grade}
v <- wheat_protein_bands()
# `v` is a decideR::grade_band_value -- a step function -- so printing it
# directly would dump its closure body and a run-specific memory address;
# the schedule itself is the breaks/values/labels triple below.
attributes(v)[c("breaks", "values", "labels")]

# A baseline protein posterior just below the milling band; a nitrogen top-up
# lifts protein, at a per-unit cost. The decision weighs the chance of clearing
# the 11.5% band against the cost.
protein <- stats::rnorm(2000L, mean = 11.0, sd = 0.5)
g_plan <- plan_grade_target(
  protein, actions = seq(0, 60, by = 10), value = v,
  quality_shift = function(rate, p) p + 0.02 * rate, cost = 0.3,
  crop = "wheat", grounding = decideR::grounding_grounded())

g_plan
```

At this cost the top-up pays for itself: the recommended input level is the
top of the candidate range, because the expected graded payoff rises
monotonically with the higher probability of clearing the 11.5% band -- the
candidate ledger below shows the full trade-off.

```{r grade-candidates}
g_plan@decision@candidates
```

## Composing a season plan

Finally, the orchestrator. `plan_season()` gathers several grain decisions into
one `grain_plan` whose grounding is the worst case across its decisions -- a plan
that rests on any unverified decision is itself unverified, the firewall applied
one level up. A plan with a declined component is flagged so a report never reads
a confident headline over it.

The nitrogen and variety decisions above come from two separate public trials,
composed here only to show the orchestrator mechanics. In a real season the
decisions would share one crop and field; the plan labels the crop explicitly so
the composition is always unambiguous.

```{r season, eval = have_agridat}
season <- plan_season(
  list(v_plan, n_plan),
  crop = "corn", season = "2026 demonstration")

season

grain_plan_table(season)
```

Every recommendation in the plan carries its grounding, and the plan as a whole
carries the worst case -- so a grower, or a downstream automation, can gate on
`grain_is_grounded()` before acting.

```{r gate, eval = have_agridat}
grain_is_grounded(season)
```

## From an upstream manifest

The examples so far assembled the yield and merit posteriors by hand. In the
orchestra, an upstream member emits an `orchestra_manifest` instead, and grainPlan
reads it directly. The manifest is read by its public S7 contract, so any object
that exposes the contract's properties works. Here is a minimal stand-in built
from a tiny S7 class, standing for the manifest a real producer would emit.

```{r manifest-mini}
mini_manifest <- S7::new_class(
  "mini_manifest",
  properties = list(
    outputs         = S7::class_any,
    metadata        = S7::class_list,
    run_id          = S7::class_character,
    emitter_package = S7::class_character))

# A yield-per-rate draws matrix, the shape a PESTO inversion emits, carried in
# `outputs` with the producer's grounding token in `metadata$grounding`.
rates_m <- seq(0, 150, by = 25)
ymax    <- stats::rnorm(1200L, 5.5, 0.3)
yld_m   <- vapply(rates_m,
                  function(r) 3 + (ymax - 3) * (1 - exp(-r / 80)),
                  numeric(1200L))
ym <- mini_manifest(outputs = yld_m,
                    metadata = list(grounding = decideR::grounding_grounded()))

plan_nitrogen_rate_from_manifest(ym, rates_m, price_grain = 350, price_n = 1.3,
                                 crop = "wheat")
```

The genomic tail reads breeding values the same way. A `breeding_values`
manifest carries point GEBVs in `outputs$gebv` alongside their genomic
uncertainty, so the reconstruction uses the genotypes' own reliability rather
than an invented spread.

```{r manifest-gebv}
gm <- mini_manifest(
  outputs  = list(gebv = c(Scepter = 0.2, Vixen = 1.1, Calibre = -0.3),
                  pev  = c(0.04, 0.04, 0.04)),
  metadata = list(grounding = decideR::grounding_grounded()))

g_dec <- plan_variety_from_manifest(gm, risk_aversion = 0.5)
g_dec
g_dec@context$gebv_sd_source
```

If a point-GEBV manifest carried no uncertainty and no `gebv_sd` were supplied,
grainPlan would stop rather than fabricate a scale for the reconstruction: a
genotype switch must not turn on a magic constant.

## Where this sits in the orchestra

grainPlan consumes the `orchestra_manifest` contract through `decideR`'s
duck-typed pipeline tail (`plan_nitrogen_rate_from_manifest()` and
`plan_variety_from_manifest()` read an upstream member's manifest directly), and
depends only on `decideR`. It reinvents nothing -- decision theory, manifest
reading, and the grounding firewall are all `decideR`'s; grainPlan adds the grain
semantics, consolidates the grain economics, and orchestrates the decisions into
a plan a practitioner can act on.

---

*Data: `lasrosas.corn` and `besag.met` from the agridat package (Wright, K.
agridat: Agricultural Datasets). The grade schedules follow published Australian
wheat and barley receival conventions; all dollar values are illustrative.*
