---
title: "From an inference result to a grain plan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From an inference result to a grain plan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1L)
```

grainPlan is the practitioner last-mile of the ORCHESTRA agricultural analytics
stack. An upstream member produces an inference result -- a yield posterior, a
multi-environment variety trial, a genomic prediction -- and grainPlan turns it
into an actionable grain-production plan: how much nitrogen to apply, which
variety to sow, whether to chase a quality grade. It is a thin grain-semantics
layer over the `decideR` decision engine: the expected-utility maximisation, the
abstention logic, and the grounding firewall are all `decideR`'s, reused whole;
grainPlan adds the grain economics and the grain meaning.

This walkthrough runs end-to-end on public data from the *agridat* package -- a
real nitrogen response trial and a real variety trial -- and shows the two
behaviours that matter for a consequential real-world tool: grounded evidence
*decides*, and unverified evidence *abstains* to the status quo.

```{r libs, message = FALSE}
library(grainPlan)
have_agridat <- requireNamespace("agridat", quietly = TRUE)
```

## A nitrogen rate from a real yield response

The *agridat* dataset `lasrosas.corn` records yield and applied nitrogen across
a precision-agriculture corn trial in Argentina (Lasrosas farm). We summarise it
into a yield-response curve, build a posterior of yield at each candidate rate,
and ask grainPlan for the profit-optimal rate.

The economics live in the grain value library. `n_rate_economics()` is the
consolidation of the profit utility the orchestra runs kept hand-building --
revenue at the grain price less the nitrogen cost -- so a profit-priced rate
decision reads as its economics rather than as an anonymous expression.

```{r nitrogen, eval = have_agridat}
data("lasrosas.corn", package = "agridat")
corn <- lasrosas.corn

# A candidate nitrogen grid spanning the trial, and a simple yield-response
# summary: mean yield in each nitrogen neighbourhood, with the trial's residual
# spread used to build a posterior of yield at each rate. This stands in for the
# yield posterior an ORCHESTRA inference member (a PESTO inversion, a TACI
# yield-response model) would emit as an orchestra_manifest.
rates <- seq(0, 120, by = 20)
fit <- stats::lm(yield ~ nitro + I(nitro^2), data = corn)
sigma <- summary(fit)$sigma

n_draws <- 1500L
yield_draws <- vapply(
  rates,
  function(r) {
    mu <- stats::predict(fit, newdata = data.frame(nitro = r))
    stats::rnorm(n_draws, mu, sigma)
  },
  numeric(n_draws))

# Grain and nitrogen prices are the caller's to supply (illustrative here).
n_plan <- plan_nitrogen_rate(
  yield_draws, rates,
  price_grain = 0.25,           # $ per unit yield (illustrative)
  price_n = 1.2,                # $ per unit nitrogen (illustrative)
  crop = "corn",
  grounding = decideR::grounding_grounded())

n_plan
```

The recommendation is a `grain_decision`: the chosen rate, the crop, a
grower-facing rationale, and -- carried beside the numbers, never inferred from
them -- the grounding token.

### The firewall: unverified evidence abstains

Now the decisive behaviour. If the yield evidence is *not* grounded -- the
upstream producer did not verify it against an external authority -- the same
decision must not be made. grainPlan inherits `decideR`'s Independent Oracle
Principle firewall: an unverified input forces an honest abstention to the
status-quo rate, never a confident wrong recommendation.

```{r firewall, eval = have_agridat}
n_plan_unverified <- plan_nitrogen_rate(
  yield_draws, rates,
  price_grain = 0.25, price_n = 1.2, crop = "corn",
  grounding = decideR::grounding_unverified())

n_plan_unverified
```

The only thing that changed is the grounding token, and it flipped *decide* into
*abstain*. This is the firewall a real grain tool needs: a nitrogen rate carries
a cost in the paddock, so an unverified yield model is held at the status quo,
not acted on.

## A variety choice from a multi-environment trial

The *agridat* dataset `besag.met` is a multi-environment variety trial: 64
genotypes across six environments. We summarise each genotype's mean performance
across environments into a posterior of merit, and ask grainPlan to rank them
under a downside-aware loss -- one that discounts a high-yielding but volatile
line against a steady one, the stability a grower wants from a sown variety.

```{r variety, eval = have_agridat}
data("besag.met", package = "agridat")
met <- besag.met

# Genotype means across environments, with the trial's environment-to-environment
# spread giving each genotype a merit posterior. (A factor-analytic MET model --
# flexyBayes::fb_met_summary -- would emit this as an orchestra_manifest; here we
# build it directly from the public trial. A few yield cells are missing, so we
# summarise with na.rm and fall back to the trial-wide spread for any genotype
# with too few records to estimate its own.)
gen_means <- tapply(met$yield, met$gen, mean, na.rm = TRUE)
gen_sd <- tapply(met$yield, met$gen, stats::sd, na.rm = TRUE)
pooled_sd <- stats::sd(met$yield, na.rm = TRUE)
gen_sd[is.na(gen_sd)] <- pooled_sd
varieties <- names(gen_means)

merit_draws <- vapply(
  seq_along(gen_means),
  function(j) stats::rnorm(1000L, gen_means[j], gen_sd[j]),
  numeric(1000L))
colnames(merit_draws) <- varieties

v_plan <- plan_variety(
  merit_draws, varieties, risk_aversion = 0.5, crop = "wheat",
  grounding = decideR::grounding_grounded())

v_plan
```

The recommendation is the leading genotype; the full ranked short-list is in the
wrapped decision's candidate ledger:

```{r shortlist, eval = have_agridat}
ledger <- v_plan@decision@candidates
ledger <- ledger[order(-ledger$expected_utility), ]
head(ledger, 5L)
```

## A quality-grade decision

Many grain payoffs are not smooth in the outcome they reward: a delivery is paid
by quality *grade*, so the value steps at a protein threshold rather than rising
continuously. grainPlan's value library consolidates these grade schedules.
`wheat_protein_bands()` prices the Australian milling-wheat grades by protein,
and `plan_grade_target()` decides how hard to push a protein-driving input under
that step payoff -- a decision governed by the posterior probability of clearing
each band, exactly the threshold economics `decideR`'s grade-band loss expresses.

```{r grade}
v <- wheat_protein_bands()
v

# A baseline protein posterior just below the milling band; a nitrogen top-up
# lifts protein, at a per-unit cost. The decision weighs the chance of clearing
# the 11.5% APW band against the cost.
protein <- stats::rnorm(2000L, mean = 11.0, sd = 0.5)
g_plan <- plan_grade_target(
  protein, actions = seq(0, 60, by = 10), value = v,
  quality_shift = function(rate, p) p + 0.02 * rate, cost = 1.3,
  crop = "wheat", grounding = decideR::grounding_grounded())

g_plan
```

## Composing a season plan

Finally, the orchestrator. `plan_season()` gathers several grain decisions into
one `grain_plan` whose grounding is the worst case across its decisions -- a plan
that rests on any unverified decision is itself unverified, the firewall applied
one level up. A plan with a declined component is flagged so a report never reads
a confident headline over it.

```{r season, eval = have_agridat}
season <- plan_season(
  list(v_plan, n_plan),
  crop = "corn", season = "2026 demonstration")

season

grain_plan_table(season)
```

Every recommendation in the plan carries its grounding, and the plan as a whole
carries the worst case -- so a grower, or a downstream automation, can gate on
`grain_is_grounded()` before acting.

```{r gate, eval = have_agridat}
grain_is_grounded(season)
```

## Where this sits in the orchestra

grainPlan consumes the `orchestra_manifest` contract through `decideR`'s
duck-typed pipeline tail (`plan_nitrogen_rate_from_manifest()` and
`plan_variety_from_manifest()` read an upstream member's manifest directly), and
depends only on `decideR`. It reinvents nothing -- decision theory, manifest
reading, and the grounding firewall are all `decideR`'s; grainPlan adds the grain
semantics, consolidates the grain economics, and orchestrates the decisions into
a plan a practitioner can act on.

---

*Data: `lasrosas.corn` and `besag.met` from the agridat package (Wright, K.
agridat: Agricultural Datasets). The grade schedules follow published Australian
wheat and barley receival conventions; all dollar values are illustrative.*
